\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-03-23 Paper/chapters/implementation-strategies.tex
table-based implementation

\begin{lstlisting}[caption={Table construction algorithm}, escapechar=|]
    tables <- [][]
    for sbox_index from 0 to 15 do
        for sbox_input from 0 to 15 do|\label{lst:tablesbox}|
            output <- sbox(sbox_input)
            output <- permute(output << (4 * sbox_index))
            tables[sbox_index][sbox_input] <- output
\end{lstlisting}

@@

\subsection{Strategies for Camellia}

>>>>

\begin{lstlisting}[caption={Table construction algorithm}, label={lst:tables}, escapechar=|]
    tables <- [][]
    for sbox_index from 0 to 15 do
        for sbox_input from 0 to 15 do|\label{lst:tablesbox}|
            output <- sbox(sbox_input)
            output <- permute(output << (4 * sbox_index))
            tables[sbox_index][sbox_input] <- output
\end{lstlisting}

@@

\section{Strategies for Camellia}

### 2023-03-23 Paper/chapters/implementation.tex
table-based implementation

We will provide and discuss implementations for the presented strategies in the
C programming language. Although directly writing Assembler code could result
in a small performance benefit, this generally increases the work necessary by
an order of magnitude for only limited results. This is why we use NEON
intrinsics.

@@

In this case, the compiler allocates vector registers for \texttt{a},
\texttt{b} and \texttt{result} and assembles the intrinsic to \texttt{MUL
Vr.8H, Va.8H, Vb.8H}. Necessary load and stores for the result and parameters
are also handled automatically. Of special interest to us are the following
intrinsics, each existing in different variants with different lane widths and
also array types: \\

@@

The implementations closely follow the aforementioned strategies;
instruction-level optimization is left to the compiler.

>>>>

We will provide and discuss implementations for the presented strategies in the
C programming language. Although directly writing Assembler code could result
in a small performance benefit, this generally increases the work necessary by
an order of magnitude for only limited results. Instruction-level optimization
and in particular register allocation is left to the compiler.

\section{Table-based}

We generate a 16 by 16 array of \texttt{uint64\_t} as explained in \ref{lst:tablesbox}.
The round function then consists of extracting each nibble, looking up the
corresponding table value and ORing this to the new cipher state.

\begin{lstlisting}[caption={Table-based round function}, language=c]
static const uint64_t table[16][16] = {
        { 0x0000000000000001UL, 0x0008000000020000UL,
          0x0000000400000000UL, 0x0008000400000000UL,
          0x0000000400020000UL, 0x0008000400020001UL,
          0x0000000000020001UL, 0x0008000000000001UL,
          0x0000000000020000UL, 0x0008000400000001UL,
          0x0008000000020001UL, 0x0000000400020001UL,
          0x0000000400000001UL, 0x0000000000000000UL,
          0x0008000000000000UL, 0x0008000400020000UL },
        // ...
        { 0x0000000010000000UL, 0x0000200000008000UL,
          0x4000000000000000UL, 0x4000000000008000UL,
          0x4000200000000000UL, 0x4000200010008000UL,
          0x0000200010000000UL, 0x0000000010008000UL,
          0x0000200000000000UL, 0x4000000010008000UL,
          0x0000200010008000UL, 0x4000200010000000UL,
          0x4000000010000000UL, 0x0000000000000000UL,
          0x0000000000008000UL, 0x4000200000008000UL }
};

uint64_t gift_64_table_subperm(const uint64_t cipher_state)
{
        uint64_t new_cipher_state = 0;

        for (size_t i = 0; i < 16; i++) {
                int nibble = (cipher_state >> (i * 4)) & 0xf;
                new_cipher_state |= table[i][nibble];
        }

        return new_cipher_state;
}
\end{lstlisting}

The round function is similar for all strategies: for each round, we extract 16
bits from the key state which are then XORed to the round key, together with
the precomputed round constants and the single bit. Key state updating is done
using masks and shifts only which gets optimized by the compiler through
bit-field extraction instructions which can extract, shift and insert a bit
field at once. Intermediate values are also optimized away.

\begin{lstlisting}[caption={Round key generation function}, language=c]
void gift_64_table_generate_round_keys(uint64_t round_keys[restrict ROUNDS_GIFT_64],
                                       const uint64_t key[restrict 2])
{
        uint64_t key_state[] = {key[0], key[1]};
        for (int round = 0; round < ROUNDS_GIFT_64; round++) {
                int v = (key_state[0] >> 0 ) & 0xffff;
                int u = (key_state[0] >> 16) & 0xffff;

                // add round key (RK=U||V)
                round_keys[round] = 0UL;
                for (size_t i = 0; i < 16; i++) {
                        int key_bit_v   = (v >> i)  & 0x1;
                        int key_bit_u   = (u >> i)  & 0x1;
                        round_keys[round] ^= (uint64_t)key_bit_v << (i * 4 + 0);
                        round_keys[round] ^= (uint64_t)key_bit_u << (i * 4 + 1);
                }

                // add single bit
                round_keys[round] ^= 1UL << 63;

                // add round constants
                round_keys[round] ^= ((round_constant[round] >> 0) & 0x1) << 3;
                round_keys[round] ^= ((round_constant[round] >> 1) & 0x1) << 7;
                round_keys[round] ^= ((round_constant[round] >> 2) & 0x1) << 11;
                round_keys[round] ^= ((round_constant[round] >> 3) & 0x1) << 15;
                round_keys[round] ^= ((round_constant[round] >> 4) & 0x1) << 19;
                round_keys[round] ^= ((round_constant[round] >> 5) & 0x1) << 23;

                // update key state
                int k0 = (key_state[0] >> 0 ) & 0xffffUL;
                int k1 = (key_state[0] >> 16) & 0xffffUL;
                k0 = (k0 >> 12) | ((k0 & 0xfff) << 4);
                k1 = (k1 >> 2 ) | ((k1 & 0x3  ) << 14);
                key_state[0] >>= 32;
                key_state[0] |= (key_state[1] & 0xffffffffUL) << 32;
                key_state[1] >>= 32;
                key_state[1] |= ((uint64_t)k0 << 32) | ((uint64_t)k1 << 48);
        }
}
\end{lstlisting}

While this approach does not rely on ARM-specific instructions, the next two
strategies do. This is why we use NEON intrinsics.

\section{Using \texttt{vperm}}

@@

In this case, the compiler allocates vector registers for \texttt{a},
\texttt{b} and \texttt{result} and assembles the intrinsic to \texttt{MUL
Vr.8H, Va.8H, Vb.8H}. Necessary loads and stores for the result and parameters
are also handled automatically. Of special interest to us are the following
intrinsics, each existing in different variants with different lane widths and
also array types: \\

@@

\section{Bitslicing}
